Certified Knowledge Compilation \\ with Application to Verified Model Counting

### 2023-05-20 doc/mc-paper.tex
Split appendix into supplement

@@

Both of these prior certification frameworks are strongly tied to the
algorithms used by the knowledge compilers and model counters.  Some
of the conditions to be checked are relevant only to specific
implementations.    Our framework is very general and is based on a small set
of proof rules.  It builds on the highly developed
concepts of clausal proof systems.  These factors were important in enabling formal verification.
In Appendix~\ref{app:comparison}
we also compare the performance of our toolchain to these two others.

@@

Appendix~\ref{app:cpog:example} shows the CPOG description for
an input formula with five clauses, yielding a POG with five
nonterminal nodes.  It explains how the clause addition and deletion
steps yield a proof of equivalence between the input formula and its POG
representation.

@@

The performance of the structural proof generator for forward implication, both in its execution time and
the size of the proof generated, can be improved by two optimizations
described here.  A key feature is that they do not require any changes
to the proof framework---they build on the power of extended
resolution to enable new logical structures.  They
involve declaring new product nodes to encode products of literals.
These nodes are not part of the POG
representation of the formula; they serve only to enable the forward
implication proof.  Here we summarize the two optimizations.
Appendix~\ref{app:optimizations} provides more details.

@@

The combination of these two optimization guarantees that i) each call
to $\validate$ for a product node will cause at most one invocation of
the SAT solver, and ii) each call to $\validate$ for any node $\nodeu$
will cause further recursive calls only once.  Our experimental
results (Appendix~\ref{app:experiments}) show that these
optimizations yield substantial benefits.

@@

We summarize the results of our experiments here.  
Appendix~\ref{app:experiments} provides a more complete description. For our
evaluation, we used the public benchmark problems from the 2022
standard and weighted model competitions.  We found that there were
180 unique CNF files among these, ranging in size from 250 to
2,753,207 clauses.
%%We ran our programs on a processor with 64~GB of
%%memory and having an attached high-speed, solid-state
%%disk.
With a runtime limit of 4000 seconds, \dfour{} completed for 124 of the
benchmark problems.  Our proof generator was able to convert all of
these into POGs, with their declarations ranging from 304 to
2,761,457,765 (median 848,784) defining clauses.
%(The
%maximum count overflowed the overflowed the 32-bit signed integers we
%used to represent clause identifiers in both the proof generator and
%checker.)
We ran our proof generator with a time limit of 10,000 seconds.
It was able to generate full proofs for 108 of the problems and
one-sided proofs for an additional 9 of them, leaving just 7 with no
verification.  The prototype checker successfully verified all of the generated
proofs.  The longest runtime for the combination of proof generator
and checker for a full proof was 13,145 seconds.
It is worth noting that, to date, we have not found
any errors in the dec-DNNF graphs generated by \dfour{}.

@@

\appendix

\section{CPOG Example}
\label{app:cpog:example}

\begin{figure}
\begin{minipage}{0.58\textwidth}
(A)  Input Formula\\[1.2ex]
\begin{tabular}{lll}
\toprule
\makebox[5mm]{ID} & \makebox[15mm]{Clauses} & \\
\midrule
1 & \texttt{-1 3 -4} & \texttt{0} \\
2 & \texttt{-1 -3 4} & \texttt{0} \\
3 & \texttt{3 -4} & \texttt{0}\\
4 & \texttt{1 -3 4} & \texttt{0} \\
5 & \texttt{-1 -2} & \texttt{0} \\
\bottomrule
\end{tabular}
\\[1.8ex]
(C) POG Declaration\\[1.2ex]
\begin{tabular}{llll}
\toprule
\makebox[5mm]{ID} & \multicolumn{2}{l}{CPOG line} & Explanation \\
\midrule
6 & \texttt{p 5 -3 -4} & \texttt{0} & $p_5 = \obar{x}_3 \pand \obar{x}_4$ \\
9 & \texttt{p 6 3 4} & \texttt{0} & $p_6 = x_3 \pand x_4$ \\
12 & \texttt{s 7 5 6 7 10} & \texttt{0} & $s_7 = p_5 \por p_6$ \\
15 & \texttt{p 8 -1 7} & \texttt{0} & $p_8 = \obar{x}_1 \pand s_7$ \\
18 & \texttt{p 9 1 -2 7} & \texttt{0} & $p_9 = x_1 \pand \obar{x}_2 \pand s_7$ \\
22 & \texttt{s 10 8 9 16 19} & \texttt{0} & $s_{10} = p_8 \por p_9$ \\
 & \texttt{r 10} && Root $r = s_{10}$\\
\bottomrule
\end{tabular}
\end{minipage}
\begin{minipage}{0.35\textwidth}
(B) POG Representation \\
\input{dd/eg4}
\end{minipage}
%% Break
\\[2.5ex]
\begin{minipage}{0.42\textwidth}
(D) Defining Clauses\\[1.2ex]
\begin{tabular}{llll}
\toprule
\makebox[5mm]{ID} & \multicolumn{2}{l}{Clauses} & Explanation \\
\midrule
\texttt{6} & \texttt{5 3 4} & \texttt{0} & Define $p_5$ \\
\texttt{7} & \texttt{-5 -3} & \texttt{0} & \\
\texttt{8} & \texttt{-5 -4} & \texttt{0} & \\
\midrule
\texttt{9} & \texttt{6 -3 -4} & \texttt{0} & Define $p_6$ \\
\texttt{10} & \texttt{-6 3} & \texttt{0} & \\
\texttt{11} & \texttt{-6 4} & \texttt{0} & \\
\midrule
\texttt{12} & \texttt{-7 5 6} & \texttt{0} & Define $s_7$ \\
\texttt{13} & \texttt{7 -5} & \texttt{0} & \\
\texttt{14} & \texttt{7 -6} & \texttt{0} & \\
\midrule
\texttt{15} & \texttt{8 1 -7} & \texttt{0} & Define $p_8$ \\
\texttt{16} & \texttt{-8 -1} & \texttt{0} & \\
\texttt{17} & \texttt{-8 7} & \texttt{0} & \\
\midrule
\texttt{18} & \texttt{9 -1 2 -7} & \texttt{0} & Define $p_9$ \\
\texttt{19} & \texttt{-9 1} & \texttt{0} & \\
\texttt{20} & \texttt{-9 -2} & \texttt{0} & \\
\texttt{21} & \texttt{-9 7} & \texttt{0} & \\
\midrule
\texttt{22} & \texttt{-10 8 9} & \texttt{0} & Define $s_{10}$ \\
\texttt{23} & \texttt{10 -8} & \texttt{0} & \\
\texttt{24} & \texttt{10 -9} & \texttt{0} & \\
\bottomrule
\end{tabular}
\end{minipage}
\begin{minipage}{0.49\textwidth}
(E) CPOG Assertions\\[1.2ex]
\begin{tabular}{llllll}
\toprule
\makebox[5mm]{ID} & \multicolumn{2}{l}{Clause} & \multicolumn{2}{l}{Hint} & Explanation \\
\midrule
\texttt{25} & \texttt{a 5 1 3} & \texttt{0} & \texttt{3 6} & \texttt{0} & $\obar{x}_1 \land \obar{x}_3 \imply p_5$ \\
\texttt{26} & \texttt{a 6 1 -3} & \texttt{0} & \texttt{4 9} & \texttt{0} & $\obar{x}_1 \land x_3 \imply p_6$ \\
\texttt{27} & \texttt{a 3 7 1} & \texttt{0} & \texttt{13 25} & \texttt{0} & $\obar{x}_3 \land \obar{x}_1 \imply s_7$  \\
\texttt{28} & \texttt{a 7 1} & \texttt{0} & \texttt{27 14 26} & \texttt{0} & $\obar{x}_1 \imply s_7$  \\
\texttt{29} & \texttt{a 8 1} & \texttt{0} & \texttt{28 15} & \texttt{0} & $\obar{x}_1 \imply p_8$  \\
\texttt{30} & \texttt{a 5 -1 3} & \texttt{0} & \texttt{1 6} & \texttt{0} & $x_1 \land \obar{x}_3 \imply p_5$ \\
\texttt{31} & \texttt{a 6 -1 -3} & \texttt{0} & \texttt{2 9} & \texttt{0} & $x_1 \land x_3 \imply p_6$ \\
\texttt{32} & \texttt{a 3 7 -1} & \texttt{0} & \texttt{13 30} & \texttt{0} & $\obar{x}_3 \land x_1 \imply s_7$  \\
\texttt{33} & \texttt{a 7 -1} & \texttt{0} & \texttt{32 14 31} & \texttt{0} & $x_1 \imply s_7$  \\
\texttt{34} & \texttt{a 9 -1} & \texttt{0} & \texttt{5 33 18} & \texttt{0} & $x_1 \imply p_9$  \\
\texttt{35} & \texttt{a 1 10} & \texttt{0} & \texttt{23 29} & \texttt{0} & $\obar{x}_1 \imply s_{10}$  \\
\texttt{36} & \texttt{a 10} & \texttt{0} & \texttt{35 24 34} & \texttt{0} & $s_{10}$ \\
\bottomrule
\end{tabular}
\\[1.5ex]
(F) Input Clause Deletions\\[1.2ex]
\begin{tabular}{lllll}
  \toprule
 \multicolumn{3}{l}{CPOG line} & Explanation\\
\midrule
 \texttt{d 1} & \texttt{36 8 10 12 16 21 22} & \texttt{0} & Delete clause 1 \\
 \texttt{d 2} & \texttt{36 7 11 12 16 21 22} & \texttt{0} & Delete clause 2 \\
 \texttt{d 3} & \texttt{36 8 10 12 17 19 22} & \texttt{0} & Delete clause 3 \\
 \texttt{d 4} & \texttt{36 7 11 12 17 19 22} & \texttt{0} & Delete clause 4 \\
 \texttt{d 5} & \texttt{36 16 20 22} & \texttt{0} &  Delete clause 5 \\
\bottomrule
\end{tabular}
\end{minipage}
\caption{Example formula (A), its POG representation (B), and its CPOG proof (C), (E), and (F)}
\label{fig:eg4:all}
\end{figure}

Figure \ref{fig:eg4:all} illustrates an example formula and shows how
the CPOG file declares its POG representation.  The input formula (A)
consists of five clauses over variables $x_1$, $x_2$, $x_3$, and
$x_4$.  The generated POG (B) has six nonterminal nodes representing
four products and two sums.  We name these by the node
type (product $\nodep$ or sum $\nodes$), subscripted by the ID of the
extension variable.
% We denote the extension variable for nodes $\nodep_i$ and $\nodes_j$ as $p_i$ and $s_j$, respectively.
  The first part of the CPOG file (C) declares
these nodes using clause IDs that increment by three or four,
depending on whether the node has two children or three.  The last two
nonzero values in the sum declarations are the hint providing the
required mutual exclusion proof.

\subsection{Node Declarations}

We step through portions of the file to provide a better understanding of the CPOG proof framework.
Figure
\ref{fig:eg4:all}(D) shows the defining clauses that are implicitly
defined by the POG operation declarations.  These do not appear in the
CPOG file.  Referring back to the declarations of the sum nodes in
Figure \ref{fig:eg4:all}(C), we can see that the declaration of node
$\nodes_7$ had clause IDs 7 and 10 as the hint.  We can see in Figure
\ref{fig:eg4:all}(A) that these two clauses form a RUP proof for the clause
$\obar{p}_5 \lor \obar{p}_6$, showing that the two children of $\nodes_7$
have disjoint models.  Similarly, node $\nodes_{10}$ is declared as having
clause IDs 16 and 19 as the hint.  These form a RUP proof for the clause
$\obar{p}_8 \lor \obar{p}_9$, showing that the two children of
$\nodes_{10}$ have disjoint models.

Figure \ref{fig:eg4:all}(E) provides the sequence of assertions
leading to unit clause 36, consisting of the literal $s_{10}$.  This clause indicates that $\nodes_{10}$ is implied by the input clauses, i.e.,
any total assignment $\assign$
satisfying the input clauses must have $\assign(s_{10}) = 1$.
Working backward, we can see that
clause 29 indicates that variable $p_8$ will be implied by the input
clauses when $\assign(x_1) = 0$, while clause 34 indicates that node $p_9$ will
be implied by the input clauses when $\assign(x_1) = 1$.  These serve as the
hint for clause 36.

Figure \ref{fig:eg4:all}(F) shows the RUP proof steps required to
delete the input clauses.  Consider the first of these, deleting
input clause $\obar{x}_1 \lor x_3 \lor \obar{x}_4$.  The requirement is to show
that there is no total assignment $\assign$ that falsifies this clause but assigns $\assign(s_{10}) = 1$.
The proof proceeds by first assuming that the clause is false, requiring
$\assign(x_1) = 1$, $\assign(x_3) = 0$, and $\assign(x_4) = 1$.  The hint then consist of unit
clauses (e.g., clause 36 asserting that $\alpha(p_{10}) = 1$) or
clauses that cause unit propagation.  Hint clauses 8 and 10 force the
assignments $\assign(p_5) = \assign(p_6) = 0$.  These, plus hint clause 12 force
$\assign(s_7) = 0$.  This, plus hint clauses 16 and 21 force $\assign(p_8) = \assign(p_9) = 0$, leading
via clause 22 to $\assign(s_{10}) = 0$.  But this contradicts clause 36,
completing the RUP proof.  The deletion hints for the other input
clauses follow similar patterns---they work from the bottom nodes of
the POG upward, showing that any total assignment that falsifies the clause
must assign $\assign(s_{10}) = 0$.

Deleting the asserted clauses is so simple that we do not show it.  It
involves simply deleting the clauses from clause number 35 down to
clause number 25, with each deletion using the same hint as were used
to add that clause.  In the end, therefore, only the defining clauses
for the POG nodes and the unit clause asserting $s_{10}$ remain,
completing a proof that the POG is logically equivalent to the input
formula.

Observe that the forward implication proof shown in Figure~\ref{fig:eg4:all}  must ``visit'' nodes
$\nodep_5$ and $\nodep_6$ twice, separately considering  assignments where $\assign(x_1) = 1$
(clauses 25 and 26) and where $\assign(x_1) = 1$ (clauses 30 and 31).
Section~\ref{app:lemma:eg} shows how to use a lemma such that these nodes are each only visited once.

\section{Optimizations for Forward Implication Proofs}

The two optimizations we have implemented exploit the power of our
general resolution framework to define new structures within the
proof.
They create new product nodes that are not part
of the POG representation.

\label{app:optimizations}

\subsection{Literal Grouping}

A single recursive step of $\validate$ can encounter product nodes
having many literals as children.  The naive formulation of $\validate$
considers each literal $\lit \in \lambda$ separately.
Literal grouping allows all literals to be validated with a single call to a SAT solver.
It collects those literals
$\lit_1, \lit_2, \ldots, \lit_m$ that cannot be validated by BCP and defines a
product node $\nodev$ having these literals as children.  The goal
then becomes to prove that any total assignment must yield 1 for extension
variable $v$.  A single call to the solver can generate this proof by invoking it on the formula
  $\simplify{\psi}{\passign} \cup \simplify{\theta_{v}}{\{ \obar{v} \}}$, which should be unsatisfiable.
  The proof steps can be mapped back into clause addition steps in the CPOG file, incorporating the
  input clauses and the defining clauses for $\nodev$ into the hints.

\subsection{Lemmas}
\label{app:lemma}

As we have noted, the recursive calling of $\validate$ starting at
root $\noder$ effectively expands the POG into a tree, and this can
lead to an exponential number of calls.
These shared subgraphs arise when the knowledge compiler employs \emph{clause caching}
to detect that the simplified set of
clauses arising from one partial assignment to the literals matches that
of a previous partial assignment~\cite{darwiche:aaai:2002}.
When this dec-DNNF node is translated into POG
node $\nodeu$, the proof generator can assume (and also check), that
there is a simplified set of clauses $\gamma_{\nodeu}$
for which the subgraph with root $\nodeu$ is its POG representation.

The proof generator can exploit the sharing of subgraphs
by constructing and proving a \emph{lemma} for each node
$\nodeu$ having $\indegree(\nodeu) > 1$.  This proof shows that any
total assignment $\assign$ that satisfies formula $\gamma_{\nodeu}$ and the defining clauses for the POG must yield
$\assign(u) = 1$.  This lemma is then invoked for every node having
$\nodeu$ as a child.
As a result, the generator will make recursive calls during a call to $\validate$ only once for each node in the POG\@.

The challenge for implementing this strategy is to find a way to
represent the clauses for the simplified formula $\gamma_{\nodeu}$ in the CPOG file.  Some may be
unaltered input clauses, and these can be used directly.  Others,
however will be clauses that do not appear in the input formula.  We
implement these by adding POG product nodes to the CPOG file to create
the appropriate clauses.  Consider an \emph{argument} clause
$C \in \gamma_{\nodeu}$ with $C = \lit_1 \lor \lit_2 \lor \cdots \lor \lit_k$.  If we
define product node $\nodev$ with arguments
$\obar{\lit}_1, \obar{\lit}_2, \ldots, \obar{\lit}_k$, then by
Table~\ref{tab:defining}(A), we will introduce a defining clause
$v \lor \lit_1 \lor \lit_2 \lor \cdots \lit_k$.  We call this a {\em
  synthetic} clause having $\obar{v}$ as the \emph{guard literal}.
That is, a partial assignment $\passign$ such that $\passign(v) = 0$ will {\em
  activate} the clause, causing it to represent argument clause $C$.  On the other
hand, a partial assignment with $\passign(v) = 1$ will
cause the clause to become a tautology and therefore have no effect.

Suppose for every clause $C_j \in \gamma_{\nodeu}$ that does not correspond to
an input clause, we generate a synthetic clause $C'_j$ with guard literal
$\obar{v}_j$, for $1 \leq j \leq m$.  Let $\gamma'_{\nodeu}$ be the formula where each clause $C_j$ is replaced by synthetic clause $C'_j$,
while input clauses in $\gamma_{\nodeu}$ are left unchanged.
Let $\lassign = \{ \obar{v}_1, \obar{v}_2, \ldots, \obar{v}_m \}$.
Invoking $\validate(\nodeu, \lassign, \gamma'_{\nodeu})$
 will then prove a lemma, given by the target clause
 $u \lor v_1 \lor v_2 \lor \cdots \lor v_m$,
 showing that any total assignment $\assign$ that activates the synthetic clauses will have $\assign(u) = 1$.

Later, when node $\nodeu$ is encountered by a call to $\validate(\nodeu, \passign, \psi)$, we invoke the lemma
by showing that each synthetic clause
$C_j$ matches some simplified clause in $\simplify{\psi}{\passign}$.  More precisely,
for $1 \leq j \leq m$,
we use clause addition to assert the clause
$\obar{v}_j \lor \bigvee_{\lit \in \passign} \obar{\lit}$,
showing that synthetic clause $C_j$ will be activated.
Combining the lemma with these activations provides a derivation of the target clause for the call to $\validate$.

Observe that the lemma structure can be hierarchical, since a shared
subgraph may contain nodes that are themselves roots of shared
subgraphs.  Nonetheless, the principles described allow the
definition, proof, and applications of a lemma for each shared node in
the graph.  For any node $\nodeu$, the first call to
$\validate(\nodeu, \passign, \psi)$ may require further recursion,
but any subsequent call can simply reuse the lemma proved by the first call.

\subsection{Lemma Example}
\label{app:lemma:eg}

\begin{figure}
(A) Additional nodes\\[1.0em]
\begin{tabular}{llll}
\toprule
\makebox[5mm]{ID} & \multicolumn{2}{l}{CPOG line} & Explanation \\
\midrule
25 & \texttt{p 11 -3 4} & \texttt{0} & $v_{11} = \obar{x}_3 \pand {x}_4$ \\
28 & \texttt{p 12  3 -4} & \texttt{0} & $v_{12} = {x}_3 \pand \obar{x}_4$ \\
\bottomrule
\end{tabular}
\\[1.0em]
(B) Implicit Clauses\\[1.2em]
\begin{tabular}{llll}
\toprule
\makebox[5mm]{ID} & \multicolumn{2}{l}{Clauses} & Explanation \\
\midrule
\texttt{25} & \texttt{11 3 -4} & \texttt{0} & Argument clause $\{x_3 ,\, \obar{x}_4\}$, activated by $\obar{v}_{11}$ \\
\texttt{26} & \texttt{-11 -3} & \texttt{0} & \\
\texttt{27} & \texttt{-11 4} & \texttt{0} & \\
\midrule
\texttt{28} & \texttt{12 -3 4} & \texttt{0} & Argument clause $\{\obar{x}_3,\,  {x}_4\}$, activated by $\obar{v}_{12}$ \\
\texttt{29} & \texttt{-12 3} & \texttt{0} & \\
\texttt{30} & \texttt{-12 -4} & \texttt{0} & \\
\bottomrule
\end{tabular}
\\[1.0em]
(C) CPOG Assertions\\[1.0em]
\begin{tabular}{llllll}
\toprule
\makebox[5mm]{ID} & \multicolumn{2}{l}{Clause} & \multicolumn{2}{l}{Hint} & Explanation \\
\midrule
\multicolumn{6}{l}{Lemma Proof} \\
\texttt{31} & \texttt{a 5 11 12 3} & \texttt{0} & \texttt{25 6} & \texttt{0} & $(\obar{v}_{11} \land \obar{v}_{12}) \land \obar{x}_3 \imply p_{5}$ \\
\texttt{32} & \texttt{a 6 11 12 -3} & \texttt{0} & \texttt{28 9} & \texttt{0} & $(\obar{v}_{11} \land \obar{v}_{12}) \land {x}_3 \imply p_{6}$ \\
\texttt{33} & \texttt{a 3 7 11 12} & \texttt{0} & \texttt{13 31} & \texttt{0} & $(\obar{v}_{11} \land \obar{v}_{12}) \land \obar{x}_3 \imply s_{7}$ \\
\texttt{34} & \texttt{a 7 11 12} & \texttt{0} & \texttt{33 14 32} & \texttt{0} & $(\obar{v}_{11} \land \obar{v}_{12}) \imply s_{7}$ \\
\midrule
\multicolumn{6}{l}{Lemma Application \#1} \\
\texttt{35} & \texttt{a -11 1} & \texttt{0} & \texttt{26 27 3} & \texttt{0} & $\obar{x}_1 \imply \obar{v}_{11}$ \\
\texttt{36} & \texttt{a -12 1} & \texttt{0} & \texttt{29 30 4} & \texttt{0} & $\obar{x}_1 \imply \obar{v}_{12}$ \\
\texttt{37} & \texttt{a 7 1} & \texttt{0} & \texttt{35 36 34} & \texttt{0} & $\obar{x}_1 \imply s_7$ \\
\midrule
\texttt{38} & \texttt{a 8 1} & \texttt{0} & \texttt{37 15} & \texttt{0} & $\obar{x}_1 \imply p_8$ \\
\midrule
\multicolumn{6}{l}{Lemma Application \#2} \\
\texttt{39} & \texttt{a -11 -1} & \texttt{0} & \texttt{26 27 1} & \texttt{0} & ${x}_1 \imply \obar{v}_{11}$ \\
\texttt{40} & \texttt{a -12 -1} & \texttt{0} & \texttt{29 30 2} & \texttt{0} & ${x}_1 \imply \obar{v}_{12}$ \\
\texttt{41} & \texttt{a 7 -1} & \texttt{0} & \texttt{39 40 34} & \texttt{0} & ${x}_1 \imply s_7$ \\
\midrule
\texttt{42} & \texttt{a 9 -1} & \texttt{0} & \texttt{5 41 18} & \texttt{0} & ${x}_1 \imply p_9$ \\
\texttt{43} & \texttt{a 1 10} & \texttt{0} & \texttt{23 38} & \texttt{0} & $\obar{x}_1 \imply s_{10}$  \\
\texttt{44} & \texttt{a 10} & \texttt{0} & \texttt{43 24 42} & \texttt{0} & $s_{10}$ \\
\bottomrule
\end{tabular}
\caption{Example of lemma definition, proof, and application}
\label{fig:eg4:lemmas}
\end{figure}

Figure~\ref{fig:eg4:lemmas} shows an alternate forward implication
proof for the example of Figure~\ref{fig:eg4:all} using a lemma to
represent the shared node $\nodes_7$.  We can see that the POG with
this node as root encodes the Boolean formula $x_3 \leftrightarrow x_4$, having a CNF representation consisting of the clauses
$\{x_3 ,\, \obar{x}_4\}$ and $\{\obar{x}_3 ,\, {x}_4\}$.  The product node
declarations shown in Figure~\ref{fig:eg4:lemmas}(A) create synthetic
clauses 25 and 28 to encode these arguments with activating literals
$\obar{v}_{11}$ and $\obar{v}_{12}$, respectively.  Clauses 31--34
then provide a proof of the lemma, stating that any assignment
$\assign$ that activates these clauses will  assign $1$ to $s_7$.
Clauses 35 and 36 state that an assignment with $\assign(x_1) = 0$
will cause the first synthetic clause to activate due to input clause
3, and it will cause the second synthetic clause to activate due to
input clause 4.  From this, clause 37 can use the lemma to state that
assigning $0$ to $x_1$ will cause $s_7$ to evaluate to $1$.  Similarly,
clauses 39 and 40 serve to activate the synthetic clauses when
$\assign(x_1) = 1$, due to input clauses 1 and 2, and clause 41 then
uses the lemma to state that assigning $1$ to $x_1$ will cause $s_7$ to
evaluate to $1$.

In this example, adding the lemma increases the proof length, but that
is only because it is such a simple formula.

\input{leanappendix}

\section{Detailed Experimental Results}
\label{app:experiments}

Our experiments are designed to evaluate the following:
\begin{itemize}
\item
  whether our toolchain can handle challenging benchmark problems,
\item
the effectiveness of some design choices and optimizations,
\item
the comparative performance of the prototype and verified checkers, and
\item
how the performance of our toolchain compares to that of the MICE model counter verifier.
\end{itemize}

\subsection{Experimental Setup}

As a test machine, we used a 2021~MacBook~Pro laptop having a 3.2~Ghz
Apple M1 processor and with 64~GB of physical memory.  We also used an
external 3~TB solid-state disk drive with an advertised read/write
speed of one gigabit per second.  Despite being a laptop, this
configuration is somewhat more powerful than the nodes typically
found in server clusters.  The fast access to storage is especially
important due to the very large files generated and processed by the
tools.  Our tools generated individual files with over 160 gigabytes
of data.

For benchmarks, we downloaded 100 files each from the 2022 standard and weighted model counting competitions:
\begin{center}
       \url{https://mccompetition.org/2022/mc_description.html}
\end{center}
We found that 20 of these were duplicates across the two tracks,
yielding a total of 180 unique benchmark problems, ranging in size
from 250 to 2,753,207 clauses.

Our default configuration for the proof generator used the structural
method for the forward implication proof, with the two optimizations
of literal grouping and lemmas.
Running with a time limit of 4000 seconds\footnote{All measured times
in this document are actual elapsed times, not CPU times.}, \dfour{}
was able to generate dec-DNNF representations for 124 of these, and it timed out on the
rest.  Our proof generator was able to convert all of the generated
dec-DNNF graphs into POGs and determine how many defining clauses they
would generate.  A POG operation with $k$ arguments requires $k+1$
defining clauses, and so the total number of defining clauses in POG $P$ equals
the number of nonterminal nodes plus the number of edges in the graph, corresponding to our definition of $|P|$ in Section~\ref{sect:pog}.
The number of defining clauses ranged from 304 to 2,761,457,765 with a median of 848,784.

\subsection{Performance of the Proof Generator and Checker}

\input{figure-runtime-d4-verify}
\input{figure-clauses-d4-verify}

For each of the 124 dec-DNNF graphs generated by \dfour{}, we ran
our proof generator and prototype checker, limiting proof
generation to 10,000 seconds.  We ran the programs to generate and check
one-sided proofs for the graphs, again with a time limit of 10,000
seconds for proof generation.

Figure~\ref{fig:d4:cpog} summarizes our results in terms of the time
required by \dfour{} (X axis) versus the sum of the times for the
proof generator and checker (Y axis).
We were able to complete a full validation for 108 of the 124
benchmarks, with times ranging from fractions of a second to 13,144
seconds, with a median of 71.6 seconds.

Relative to the runtime for \dfour{}, two problems ran faster with the
generator and checker.  These were for problems having small numbers
of models (relative to the number of variables), and so most of the
time spent by both \dfour{} and our proof generator was in running a
CDCL solver to search the very sparse solution space.  \cadical{}
generally outperforms the miniSAT-based solver used by \dfour{}.  At
the other extreme, one problem that required only 19 seconds for
\dfour{} required 8657 seconds to generate a proof and 45 seconds to
check it.  This benchmark appears to stem from weak performance
by our implementation of BCP\@.  Overall the ratios
between the combined generation plus checking times versus the time
for \dfour{} had a harmonic mean of 5.5.

Of the 16 benchmarks that could not be fully validated, one had so
many defining clauses that it overflowed the 32-bit signed integers
our programs use for clause identifiers.  Another exited due to the
virtual memory limit imposed by the operating system, and the other 14
timed out.  Of the 15 that did not overflow the clause counter, we
were able to complete a one-sided verification for 9, but the other 6
timed out.  Overall, we were able to provide some level of
verification for 94\% of the benchmark problems.

Figure~\ref{fig:defining:total} compares the number of defining
clauses (X axis) with the total number of clauses (Y axis) for the 108
problems that were fully verified.  These totals include the defining
clauses for the POG, the additional defining clauses introduced to
support literal grouping and lemmas, and the clauses added by RUP
steps in the forward implication proof.  These totals ranged from 554
to 770,382,773, with a median of 1,719,245.  The ratios of the total
clauses versus defining clauses ranged from 1.54 to 6073, with a
harmonic mean of 3.13.  The high numbers were for problems having very
few models, and so the proof clauses must encode large proofs of
unsatisfiability.

\subsection{Monolithic Forward Implication Proofs}
\label{app:experiment:monolithic}

To assess the relative merits of the two approaches to forward
implication proof generation, we ran the proof generator in monolithic
mode on the 92 problems for which the combination of structural proof generation
and checking required at most 3000 seconds and then kept only those
measurements for which at least one of the two approaches had combined
times (generation plus checking) of at most 1000 seconds.  There were
a total of 83 such problems.  Figures~\ref{fig:monolithic:time} and
\ref{fig:monolithic:clauses} show comparisons between the two
approaches in terms of time and total clauses.

Examining Figure~\ref{fig:monolithic:time}, we can see that the
monolithic approach ran faster than the structural approach for 23 of
the problems, including a case where the generator and checker
required only one second, versus 71 seconds for the structural
approach.  On the other hand, the monolithic approach timed out for 13
of the cases and was slower for the other 37.  In general, we can see
the monolithic approach faring worse on larger problems.

Figure~\ref{fig:monolithic:clauses} shows the total number of clauses
for the two approaches for the 70 cases where both approaches completed within 1000 seconds.
As can be seen, the monolithic approach consistently produces
shorter proofs, perhaps due to the benefits of the proof trimming by
\dtrim{}.

\input{figure-runtime-monolithic}
\input{figure-clauses-monolithic}

These experiments suggest that a hybrid of the two approaches might
yield the best results in terms of consistency, runtime, and proof
size.  It would begin the recursive descent of $\validate$ according
to a structural approach, but monolithically generate
the proof once it encounters a sufficiently small subgraph.

\subsection{Impact of Optimizations}
\label{app:experiment:optimize}

\input{figure-clauses-optimizations}

To assess the impact of the two optimizations: literal grouping and
lemmas, we ran the proof generator on the 80 problems for which the
combination of the two optimizations yielded proofs totaling at most
10~million clauses.  Compared to the problems used to assess monolithic
proof generation (Section~\ref{app:experiment:monolithic}), this set included one that had a short proof but
long runtime and excluded four that had short runtimes but long
proofs.  We then ran the proof generator on all three combinations of
the optimizations being partially or totally disabled.
Figure~\ref{fig:optimized:lessoptimized} shows how disabling these optimizations affected the
total number of clauses in the proofs, where the X axis indicates the
total number with both optimizations enabled, while the Y axis
indicates the number with one (left) or both (right) optimizations
disabled.

As the figures illustrated, each optimization on its own can shorten
the proofs, sometimes substantially, and the two combine to have even
greater effect.  Of the 80 benchmark problems evaluated, and comparing
with both optimizations disabled, 47 had the number of proof clauses
reduced by a factor of 2 by using lemmas, 14 by using literal merging,
and 60 by enabling both optimizations.  In the extreme cases, a lack
of lemmas caused one proof to grow by a factor of 52.5, while a lack
of literal grouping causes another proof to grow by a factor of 39.6.

The time performance of these optimizations is less dramatic, but
still significant.  Lemmas sometimes made proof generation slower, but
in one case ran $38\times$ faster.  Similarly, literal grouping could
take longer, but in one case ran $3.9\times$ faster.

Overall, it is clear that these two optimizations are worthwhile, and
sometimes critical, to success for some benchmarks.

\subsection{Certifying Preprocessed Formulas}

\input{figure-runtime-preprocess}
\input{figure-clauses-preprocess}

The \dfour{} knowledge compiler can optionally perform three different
forms of preprocessing on the input formula.  These transform the
formula while preserving logical equivalence~\cite{lagniez:aaai:2014}.
Proving that the combination of preprocessing plus knowledge
compilation preserved logical equivalence would be a valuable
capability.  However, proof generators that operate based on the
structure of the input clauses, including our structural approach,
will generally fail in this case, since the compilation was not based
on the original input clauses.

>>>>

We have developed an online supplement to this
paper~\cite{bryant:sat:2023:supplement} containing additional
information, including a worked example, details on the algorithms,
and extensive experimental results.  The archive also includes copies
of the programs and a set of benchmark files for testing.

@@

Both of these prior certification frameworks are strongly tied to the
algorithms used by the knowledge compilers and model counters.  Some
of the conditions to be checked are relevant only to specific
implementations.    Our framework is very general and is based on a small set
of proof rules.  It builds on the highly developed
concepts of clausal proof systems.  These factors were important in enabling formal verification.
In the supplement~\cite{bryant:sat:2023:supplement},
we also compare the performance of our toolchain to these two others.

@@

The supplement~\cite{bryant:sat:2023:supplement},
shows the CPOG description for
an input formula with five clauses, yielding a POG with five
nonterminal nodes.  It explains how the clause addition and deletion
steps yield a proof of equivalence between the input formula and its POG
representation.

@@

The performance of the structural proof generator for forward implication, both in its execution time and
the size of the proof generated, can be improved by two optimizations
described here.  A key feature is that they do not require any changes
to the proof framework---they build on the power of extended
resolution to enable new logical structures.  They
involve declaring new product nodes to encode products of literals.
These nodes are not part of the POG
representation of the formula; they serve only to enable the forward
implication proof.  Here we summarize the two optimizations.
The supplement~\cite{bryant:sat:2023:supplement} provides more details.

@@

The combination of these two optimization guarantees that i) each call
to $\validate$ for a product node will cause at most one invocation of
the SAT solver, and ii) each call to $\validate$ for any node $\nodeu$
will cause further recursive calls only once.  Our experimental
results~\cite{bryant:sat:2023:supplement} show that these
optimizations yield substantial benefits.

@@

We summarize the results of our experiments here.  
The supplement~\cite{bryant:sat:2023:supplement}
provides a more complete description. For our
evaluation, we used the public benchmark problems from the 2022
standard and weighted model competitions.  We found that there were
180 unique CNF files among these, ranging in size from 250 to
2,753,207 clauses.
%%We ran our programs on a processor with 64~GB of
%%memory and having an attached high-speed, solid-state
%%disk.
With a runtime limit of 4000 seconds, \dfour{} completed for 124 of the
benchmark problems.  Our proof generator was able to convert all of
these into POGs, with their declarations ranging from 304 to
2,761,457,765 (median 848,784) defining clauses.
%(The
%maximum count overflowed the overflowed the 32-bit signed integers we
%used to represent clause identifiers in both the proof generator and
%checker.)
We ran our proof generator with a time limit of 10,000 seconds.
It was able to generate full proofs for 108 of the problems and
one-sided proofs for an additional 9 of them, leaving just 7 with no
verification.  The prototype checker successfully verified all of the generated
proofs.  The longest runtime for the combination of proof generator
and checker for a full proof was 13,145 seconds.
It is worth noting that, to date, we have not found
any errors in the dec-DNNF graphs generated by \dfour{}.

@@
